---
title: "Data Processing"
---

## Data Acquisition

Describe where you got the data and how you downloaded it.


1. Zillow Home Value & Rent Data
Source: Zillow Research Data Portal (https://www.zillow.com/research/data/)

Files Downloaded:

Metro_zhvi_uc_sfrcondo_tier.csv - All Homes Time Series, Metro Level

Metro_zori_sm.csv - Observed Rent Index, Metro Level

Acquisition Method: Direct download via web browser. Files contain monthly values for all metropolitan statistical areas.



2. Census Income Data
Source: U.S. Census Bureau American Community Survey (ACS)

Table Used: S1903 - Median Income in the Past 12 Months

Geography: Metropolitan Statistical Areas

Years: 2014-2022 (most recent available)


3. Mortgage Rate Data
Source: Federal Reserve Economic Data (FRED)

Series: MORTGAGE30US - 30-Year Fixed Rate Mortgage Average

Acquisition Method: Downloaded via the fredr R package interface

Time Period: Monthly averages, 2014-2025


```{r, eval=FALSE}
library(readr)
Census_Income_Data <- read_csv("ACSST1Y2024.S1903-2025-11-19T223915.csv")
Metropolitan_level_Zillow_Home_Value_Index <- read_csv("Metro_zhvi_uc_sfrcondo_tier_0.33_0.67_sm_sa_month.csv")
Zillow_Observed_Rent_Index <- read_csv("Metro_zori_uc_sfrcondomfr_sm_month.csv")
Mortage_interest_rates <- read_csv("MORTGAGE30US.csv")
```


# Data Cleaning

Show the steps you took to clean the data.

```{r}

```

# Exploratory Data Analysis

This is where you put all your exploratory plots (the ones that didn't make it into the final report).

```{r}
#| echo: true
# Exploratory histogram
# ggplot(cleaned_data, aes(x = my_variable)) + geom_histogram()
```
